---
title: "Class 8: Breast Cancer Mini Project"
author: "Angelika (PID: A17475228)"
format: pdf
toc: TRUE
---

## Background

In today's class we will be employing all the R techniques for data analysis that we have learned thus far- including the machine learning methods of clustering and PCA- to analyze real breast cancer biopsy data.

## Data Import

The data is in CSV format:

```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
```

We peak at the data

```{r}
head(wisc.df, 3)
```

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```

> Q3. How many variables /features in the data are suffixed with _mean?

```{r}
length (grep("_mean", colnames(wisc.df)))
```
```{r}
colnames(wisc.df)
```
We need to remove the `diagnosis` column before we do any further analysis of this dataset- we don't want to pass this to PCA, etc. We will save it as a separate wee vector that we can use for later to compare our findings to those of experts.

```{r}
wisc.data <- wisc.df[ ,-1]
diagnosis <- wisc.df$diagnosis
```

## Principle Component Analysis (PCA)

The main function in base R is called `prcomp()`. We will use the optional argument `scale=TRUE` here at the data columns/features/dimensions are on very different scales in the original data.

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
```

```{r}
attributes(wisc.pr)
```

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

```{r}
summary(wisc.pr)
```

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
pve <- pr.var / sum(pr.var)
```

```{r}
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation ["concave.points_mean", "PC1"]
```


## Hierarchical Clustering

The goal of this section is to do hierarchical clustering of the original datato see if there is any obvious grouping into malignant and benign clusters.

In short, these results are not good!

First, we will scale our `wisc.data` then calculate a distance matrix, then pass to `hclust()`:

```{r}
wisc.dist <- dist ( scale(wisc.data))
wisc.hclust <- hclust(wisc.dist)
plot(wisc.hclust)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=2)
table(wisc.hclust.clusters)
```

## Combining Methods

The idea here is that I can take my new variables (i.e. the scores in the PCs `wisc.pr$x) that are better descriptors of the data set than the original features (i.e. the 30 columns ub `wisc.data`) and use these as a basis for clustering.

```{r}
pc.dist <- dist(wisc.pr$x[ ,1:3])
wisc.pr.hclust <- hclust(pc.dist, method= "ward.D2")
plot(wisc.pr.hclust)
```

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

I can now run `table()` with both clustering `grps` and the expert `diagnosis`

```{r}
table(grps, diagnosis)
```

Our cluster "1" has 179 "M" diagnosis or cluster "2" has 333 "B" diagnosis

179 TP
24 FP
333 TN
33 FN

Sensitivity: TP/(TP+FN)

```{r}
179/(179+33)
```

Specificity: TN/(TN+FP)

```{r}
333/(333+24)
```

## Prediction

We will use the `predict()` function that will take our PCA model from before and new cancer cell data and project that data onto our PCA space.

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

