---
title: "Class 17: Analyzing sequencing data in the cloud"
author: "Angelika (PID: A17475228)"
format: pdf
toc: true
---

## Downstream analysis

Back on our laptop we can now use R and Bioconductor tools to further explore this large scale data-set.

For example there is an R function called tximport() in the tximport package, which enables straightforward import of Kallisto results.

With each sample having its own directory containing the Kallisto output, we can import the transcript count estimates into R using: `BiocManager::install("tximport")`

library(tximport)

```{r}
library(tximport)

# setup the folder and file-names to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```
**We now have our estimated transcript counts for each sample in R. We can see how many transcripts we have for each sample:**

```{r}
colSums(txi.kallisto$counts)
```

**And how many transcripts are detected in at least one sample:**

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

**Before subsequent analysis, we might want to filter out those annotated transcripts with no reads:**

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

**And those with no change over the samples:**

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## Principal Component Analysis

We can now apply any exploratory analysis technique to this counts matrix. As an example, we will perform a PCA of the transcriptomic profiles of these samples.

Now we compute the principal components, centering and scaling each transcript’s measured levels so that each feature contributes equally to the PCA:

```{r}
pca <- prcomp(t(x), scale=TRUE)

summary(pca)
```
**Now we can use the first two principal components as a co-ordinate system for visualizing the summarized transcriptomic profiles of each sample:**

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

> Q. Use ggplot to make a similar figure of PC1 vs PC2 and a separate figure PC1 vs PC3 and PC2 vs PC3.

```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","pink","red","purple")

ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()
```
**PC1 vs PC3**
```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","pink","red","purple")

ggplot(pca$x) +
  aes(PC1, PC3, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()
```
**PC2 vs PC3**
```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","pink","red","purple")

ggplot(pca$x) +
  aes(PC2, PC3, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()
```

## Optional: Differential-expression analysis

We can use DESeq2 to complete the differential-expression analysis that we are already familiar with:

An example of creating a DESeqDataSet for use with DESeq2:

```{r}
library(DESeq2)

sampleTable <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(sampleTable) <- colnames(txi.kallisto$counts)

dds <- DESeqDataSetFromTximport(txi.kallisto,
                                sampleTable, 
                                ~condition)
```

**Using counts and average transcript lengths from tximport**

```{r}
dds <- DESeq(dds)

res <- results(dds)
head(res)
```

